\textbf{\sc Towards Deliberative Control in Marine Robotics}

### 2012-02-24 trunk/europa.tex
final updates

\begin{figure}
\centering
\includegraphics[scale=0.4]{figs/EUROPA-Architecture.pdf}
\caption{\small The \eu software architecture.}
\label{fig:europa-architecture}
\end{figure}

Fig. \ref{fig:europa-architecture} shows the main architectural
components in \eu and their relationships. The \textbf{Constraint
  Reasoning Engine} (\texttt{CRE}) manages domain types, variables and
constraints that define relationships among them. It also provides an
efficient arc consistency mechanism \cite{mackworth77}. The
\texttt{CRE} is designed so that specialized reasoning algorithms for
new constraints can be easily and efficiently plugged in. The
\textbf{Plan Database} (\texttt{PDB}) manages object and token
types. It also maintains the state of partial plans in terms of
object, variable and token instances. This constitutes the data
foundation on which end user applications and automated planners
within \eu are built. The \textbf{Rules Engine} (\texttt{RE})
modularly adds the semantics of domain rules that describe
dependencies between tokens. The \textbf{Solver Module} provides the
framework for flaw detection and resolution that is needed to
implement search algorithms for planning. The \textbf{API Layer}
provides access to services from all the different modules in a
programmatic way so that \eu can be easily embedded into specialized
applications.

@@

\item[\textbf{Variable Types and Instances}:] All object attributes
  are represented as variables in this model. For instance, it uses
  integer variables to represent the built-in temporal parameters for
  each token (activity duration, start and end times), string
  variables to represent \textit{Location} and \textit{Product} names,
  and object reference variables to represent the structural
  relationships between different objects (for instance, each Agent
  maintains a variable to keep track of its \textit{AgentLocation}).

@@

\item[\textbf{Action Types}:] In our example, \textit{Agent} is solely
  the class that has behavior associated with it. Consequently, it
  defines action types to represent when the agent is taking action to
  \texttt{Go} to a \textit{Location}, or to \texttt{Buy} a
  \textit{Product}. % We'll see how the actual behavior for those
  % actions is specified below.

@@

\item[\textbf{Subgoals}:] In \texttt{Agent::Go}, subgoals are used to
  specify that the Shopping Agent's location must transition from
  origin to destination in a temporally consistent way. The
  \texttt{[meets, equals, metby]} operators (see Section
  \ref{sec:europa:inference}) are used to specify subgoals, along with
  the temporal relationship each token representing a subgoal has with
  the the token representing the \texttt{Agent::Go} action. Also,
  \texttt{[condition, effect]} annotations can be used when defining a
  subgoal inside an action. Currently \eu does not associate semantics
  with these annotation; but they can be extremely useful for search
  modules when creating plans.

@@

\label{eu:model}
\end{verbatim}

This problem instance consists of \textbf{Variables} and
\textbf{Objects}, \textbf{Facts}, \textbf{Goals} and
\textbf{Constraints}. In our example we define temporal variables that
are used to specify the planning horizon (\texttt{horizonStart} and
\texttt{horizonEnd}) and objects are used to specify locations,
products, product-location associations and the Shopping Agent that we
want to plan for. Facts are used to specify the current state of the
world and do not need any support and are represented as tokens in the
\texttt{PDB}. In this case we use facts to specify that the Shipping
Agent is initially at home. Goals are also represented as tokens in
the \texttt{PDB}. Different from facts, goals trigger inference in \eu
which determines whether flaws (unresolved subgoals, overloaded
resources, etc.) need to be resolved; if so, search modules can be
invoked to resolve those flaws as part of the deliberation
process. Finally, constraints are restrictions on and between actions
and goals. In the example above, we use temporal constraints to
specify when facts hold and when the goals should be
accomplished. % As
% always, all constraints in \eus library could be used.

\begin{figure}[!htb]
\centering
\includegraphics[scale=0.3]{figs/Allen-algebra.pdf}
\caption{\small Temporal relations defined within the planner are
  based on \texttt{Allen Algebra} \cite{allen84} relations shown
  above.}
\label{fig:allen-algebra}
\end{figure}

In \eu, problem domains use \texttt{Allen Algebra} \cite{allen84}
temporal relations shown in Fig. \ref{fig:allen-algebra} which allow
encoding of temporal constraint relationships between tokens.
\kcomment{This para could do with a little embellishment especially
  since it is the only material that actually talks about the kinds of
  constraints and their origin.JB: I don't really know what else to
  say given the table above.}

@@

Resource envelopes can be useful when trying to evaluate plan
brittleness in the face of changing activity durations or start
times. They can also be useful when searching for plans; however they
are relatively expensive to compute. Therefore their ability to prune
the search space must be weighted against the ability to search faster
\cite{6007772}.

@@

Given the above, an automated problem solver can be created using, a
\emph{domain model} that describes the variable, object, predicate and
action types that are relevant for the problem, a \emph{problem
  instance} or initial state that consists of variable and object
instances that exist for the entire planning horizon, \emph{temporally
  scoped predicate and action} instances and \emph{temporally scoped
  goals}. This information is retained in \eus \texttt{PDB} so that
inference and search mechanisms can be used to look for a problem
solution.

Since temporal intervals may extend indefinitely in both positive and
negative directions, for problems with a temporal dimension it is also
common to specify a \emph{planning horizon} when invoking a solver;
any decisions that fall completely outside of the horizon can be
ignored. This is particularly useful for recurrent activities;
consider for instance a model where a medical checkup has to occur
every $3$ months. The user can model a recurring activity and then use
the horizon to ensure that a finite number of medical checkup
activities are generated as part of a plan. This is also useful to
ignore activities that may occur far in the future or in the past to
be relevant for a particular plan, ensuring that the solver is making
as few commitments as necessary and making planning tractable.

@@

\item[\textbf{Open Condition}:] an open condition is an
  \texttt{INACTIVE} token. \texttt{INACTIVE} tokens can be generated
  by explicitly posted goals, or when a token is activated, rules may
  activate the creation of \texttt{INACTIVE} slave tokens as described
  in Section \ref{sec:europa:pr}. Open Conditions can be resolved in
  three ways:

@@

  \item[\textbf{Merging}:] A free activity \kcomment{this could be
      defined, since no def. for 'free' is provided} is merged with a
    matching activity already in the plan. An activity $a$ is said to
    match an activity $a?$ iff $a$ and $a?$ unify \cite{russelnorvig}
    and the temporal constraints involving $a$ are satisfied by
    $a?$. Thus, $a$ and $a?$ can be considered the same activity
    without the need to introduce $a$ in the plan. Consequently, the
    domain rules associated with $a$ are not introduced into the
    planning process, since they are already triggered when $a?$ was
    introduced in the plan.

  \item[\textbf{Activation}:] A new activity $a$ is introduced in the
    current plan associating it with the appropriate timeline, without
    choosing a specific time slot; if its timing causes a problem a
    Threat flaw will be triggered (see below).  The domain rules
    associated with $a$ are applied and the subgoal activities
    resulting from those domain rules are introduced as inactive
    tokens.  This results in a number of open condition flaws,
    corresponding to the new subgoal activities.

@@

Algorithm \ref{alg:europa:solve} provides for a sound and complete
search, assuming that no flaws or available refinement operators are
pruned unnecessarily.  Dead-ends in the search are discovered through
constraint propagation, a vehicle for evaluating the consistency of a
partial plan and also for filtering infeasible values from
consideration prior to commitment. % This allows in some cases a strong
% look-ahead capability which is essential for tractable
% search.
Consistency testing is initiated by the \texttt{isInconsistent}
procedure. The algorithm permits a heuristically controlled search by
applying orderings for \texttt{chooseFlaw} and
\texttt{makeDecisionPoint} resulting in a chronologically-backtracked
search. Heuristics can be implemented by applying ordering/filtering
policies for \texttt{chooseFlaw} and \texttt{makeDecisionPoint}, for
instance, by dealing with earliest flaws first, or by making planning
decisions (open condition flaws) before scheduling decision (threats).

It should be emphasized that while this algorithm is commonly employed
with \eu implementations, it is only one of many that could be
implemented. As a simple example, consider the following solver for
the $N$-Queens problem, where an integer variable V$_i$ represents the
queen in column $i$; V$_1$ $= 3$ implies that the variable on column
$1$ is placed on row $3$.  Non-attack constraints are posted as
arithmetic constraints among the V$_i$ variables to ensure that all
the rows and diagonals contain only one queen. \eus Constraint
Reasoning Engine or \texttt{CRE} can then be used to compute the
non-attack constraint violations, with \texttt{ce.getViolations()} a
call on the API, to pick the queen that has he most violations.  By
implementing \texttt{getQueenWithMaxViolation()} and looking ahead to
see what moves are possible for a queen that would remove its
violations with the implementing of \texttt{getMoves()}, this solver
provides a viable planner implementation for a range of possible
initial states.

\begin{algorithm}[H]
  \KwIn{ConstraintEngine $ce$}
  \KwIn{int $maxIter$}
  \Begin{
    int $curIter \leftarrow 0$ \;
    \While{$\left( ce.getViolations() > 0 \right) ~ \& ~ \left( curIter <
        maxIter \right)$ }{
      Variable $queenToMove \leftarrow getQueenWithMaxViolation(ce)$
      \;
      SortedSet$<$int$> moves \leftarrow getMoves(ce, queenToMove,
      curPos)$ \;
      bool $moved \leftarrow $ \texttt{false} \;
      \ForEach{int $newPos$ of $moves$}{
        $moved \leftarrow tryMove(queenToMove, curPos, newPos)$\;
        \lIf{$moved$}{ \textbf{break} \; }
      }
      \lIf{$! moved$}{
        $forceMove(queenToMove, curPos, moves.first())$ \;
      }
      $checkSolution()$ \tcp*{See if we have a new best solution}
      $curIter$\texttt{++} \;
    }
  }
  \caption{$solveNQueens(ce, maxIter)$}
\end{algorithm}

>>>>

\begin{figure}[b]
\centering
\includegraphics[scale=0.4]{figs/EUROPA-Architecture.pdf}
\caption{\small The \eu software architecture.}
\label{fig:europa-architecture}
\end{figure}

Fig. \ref{fig:europa-architecture} shows the main architectural
components in \eu and their relationships. The \textbf{Constraint
  Reasoning Engine} (\texttt{CRE}) manages domain types, variables and
constraints that define relationships among them. It also provides an
efficient arc consistency mechanism \cite{mackworth77}. The
\texttt{CRE} is designed so that specialized reasoning algorithms for
new constraints can be easily and efficiently plugged in. The
\textbf{Plan Database} (\texttt{PDB}) manages object and token
types. It also maintains the state of partial plans in terms of
object, variable and token instances. This constitutes the data
foundation on which end user applications and automated problem solvers
 are built. The \textbf{Rules Engine} (\texttt{RE})
modularly adds the semantics of domain rules that describe
dependencies between tokens. The \textbf{Solver Module} provides the
framework for flaw detection and resolution that is needed to
implement search algorithms for planning. The \textbf{API Layer}
provides access to services from all the different modules in a
programmatic way so that \eu can be easily embedded into specialized
applications.

@@

\item[\textbf{Variable Types and Instances}:] All object attributes
  are represented as variables in this model. For instance, it uses
  integer variables to represent the built-in temporal parameters for
  each token (duration, start and end times), string
  variables to represent \textit{Location} and \textit{Product} names,
  and object reference variables to represent the structural
  relationships between different objects (for instance, each Agent
  maintains a variable to keep track of its \textit{AgentLocation}).

@@

\item[\textbf{Action Types}:] In our example, \textit{Agent} is solely
  the class that has behavior associated with it. Consequently, it
  defines action types to represent when the agent is taking action to
  \texttt{Go} to a \textit{Location}, or to \texttt{Buy} a
  \textit{Product}.

@@

\item[\textbf{Subgoals}:] In \texttt{Agent::Go}, subgoals are used to
  specify that the Shopping Agent's location must transition from
  origin to destination in a temporally consistent way. The
  \texttt{[meets, equals, metby]} operators (see Section
  \ref{sec:europa:inference}) are used to specify subgoals, along with
  the temporal relationship each token representing a subgoal has with
  the the token representing the \texttt{Agent::Go} action. Also,
  \texttt{[condition, effect]} annotations can be used when defining a
  subgoal inside an action. \eu does not intrinsicaly associate semantics
  with these annotations; but they can be extremely useful for search
  modules when creating plans.

@@

\end{verbatim}

This problem instance consists of \textbf{Variables} and
\textbf{Objects}, \textbf{Facts}, \textbf{Goals} and
\textbf{Constraints}. In our example we define temporal variables that
are used to specify the planning horizon (\texttt{horizonStart} and
\texttt{horizonEnd}) and objects are used to specify locations,
products, product-location associations and the Shopping Agent that we
want to plan for. Facts are used to specify the current state of the
world, do not need any support and are represented as tokens in the
\texttt{PDB}. In this case we use facts to specify that the Shipping
Agent is initially at home. Goals are also represented as tokens in
the \texttt{PDB}. Different from facts, goals trigger inference in \eu
which determines whether flaws (unresolved subgoals, overloaded
resources, etc.) need to be resolved; if so, search modules can be
invoked to resolve those flaws as part of the deliberation
process. Finally, constraints are restrictions on and between facts
and goals. In the example above, we use temporal constraints to
specify when facts hold and when the goals should be accomplished. In
the model above, \texttt{before} and \texttt{after} are \texttt{Allen
  Algebra} \cite{allen84} operators, which \eu supports to specify
temporal relations between tokens (specifically between their start
and end timepoints). The algebra is particularly effective in dealing
with interval arithmetic; its elements are summarized in
Fig. \ref{fig:allen-algebra}

\begin{figure}[!t]
\centering
\includegraphics[scale=0.4]{figs/Allen-algebra.pdf}
\caption{\small Temporal relations defined within the planner are
  based on \texttt{Allen Algebra} \cite{allen84} relations shown
  above.}
\label{fig:allen-algebra}
\end{figure}

@@

Resource envelopes can be useful when trying to evaluate plan
brittleness in the face of changing activity durations or start
times. They can also be useful when searching for plans; however they
are relatively expensive to compute. Therefore their ability to prune
the search space must be weighted against the ability to search faster
\cite{6007772, morris11}.

@@

Given the above, an automated problem solver can be created using a
\emph{domain model} that describes the variable, object, predicate and
action types that are relevant for the problem, a \emph{problem
  instance} or initial state that consists of variable and object
instances that exist for the entire planning horizon, \emph{temporally
  scoped predicate and action} instances and \emph{temporally scoped
  goals}. This information is retained in \eus \texttt{PDB} so that
inference and search mechanisms can be used to look for a problem
solution.

Since temporal intervals may extend indefinitely in both positive and
negative directions, for problems with a temporal dimension it is also
common to specify a \emph{planning horizon} when invoking a solver;
any decisions that fall completely outside of the horizon can be
ignored. This is useful to ignore activities that may occur far in the
future or in the past to be relevant for a particular plan, ensuring
that the solver is making as few commitments as necessary and reducing
the complexity of plan generation.  The \emph{planning horizon} is
also useful when dealing with recurrent activities; consider for
instance a model where a medical checkup has to occur every $3$
months. The user can model a recurring activity and then use the
horizon to ensure that a finite number of medical checkup activities
are generated as part of a plan.

@@

\item[\textbf{Open Condition}:] an open condition is an
  \texttt{INACTIVE} token. \texttt{INACTIVE} tokens can be generated
  by explicitly posted goals, or when a token is activated, rules may
  cause the creation of \texttt{INACTIVE} slave tokens as described
  in Section \ref{sec:europa:pr}. Open Conditions can be resolved in
  three ways:

@@

  \item[\textbf{Merging}:] An \texttt{INACTIVE} token is merged with a
    matching \texttt{ACTIVE} token already in the plan. A token $a$ is said to
    match another token $a?$ iff $a$ and $a?$ unify \cite{russelnorvig}
    and the temporal constraints involving $a$ are satisfied by
    $a?$. Thus, $a$ and $a?$ can be considered the same token
    without the need to introduce $a$ in the plan. Consequently, the
    domain rules associated with $a$ are not introduced into the
    planning process, since they are already triggered when $a?$ was
    introduced in the plan.

  \item[\textbf{Activation}:] A token $a$ is introduced in the
    current plan associating it with the appropriate timeline, without
    choosing a specific time slot; if its timing causes a problem a
    Threat flaw will be triggered (see below).  The domain rules
    associated with $a$ are applied and the subgoal activities
    resulting from those domain rules are introduced as \texttt{INACTIVE}
    tokens.  This results in a number of open condition flaws,
    corresponding to the new subgoal activities.

@@

Algorithm \ref{alg:europa:solve} provides for a sound and complete
search, assuming that no flaws or available refinement operators are
pruned unnecessarily.  Dead-ends in the search are discovered through
constraint propagation, a vehicle for evaluating the consistency of a
partial plan and also for filtering infeasible values from
consideration prior to commitment. Consistency testing is initiated by
the \texttt{isInconsistent} procedure.

The algorithm permits a heuristically controlled search by
applying orderings for \texttt{chooseFlaw} and
\texttt{makeDecisionPoint} resulting in a chronologically-backtracked
search. Heuristics can be implemented by applying ordering/filtering
policies for \texttt{chooseFlaw} and \texttt{makeDecisionPoint}, for
instance, by dealing with earliest flaws first, or by making planning
decisions (open condition flaws) before scheduling decision (threats).

It should be emphasized that while this algorithm is commonly employed
by applications that embed  \eu, it is only one of many that could be
implemented. As a simple example, consider the following solver for
the $N$-Queens problem, where an integer variable V$_i$ represents the
queen in column $i$; V$_1$ $= 3$ implies that the variable on column
$1$ is placed on row $3$.  Non-attack constraints are posted as
arithmetic constraints among the V$_i$ variables to ensure that all
the rows and diagonals contain only one queen.

\begin{algorithm}[H]
  \KwIn{ConstraintEngine $ce$}
  \KwIn{int $maxIter$}
  \Begin{
    int $curIter \leftarrow 0$ \;
    \While{$\left( ce.getViolations() > 0 \right) ~ \& ~ \left( curIter <
        maxIter \right)$ }{
      Variable $queenToMove \leftarrow getQueenWithMaxViolation(ce)$
      \;
      SortedSet$<$int$> moves \leftarrow getMoves(ce, queenToMove,
      curPos)$ \;
      bool $moved \leftarrow $ \texttt{false} \;
      \ForEach{int $newPos$ of $moves$}{
        $moved \leftarrow tryMove(queenToMove, curPos, newPos)$\;
        \lIf{$moved$}{ \textbf{break} \; }
      }
      \lIf{$! moved$}{
        $forceMove(queenToMove, curPos, moves.first())$ \;
      }
      $checkSolution()$ \tcp*{Evaluate if a new best solution}
      $curIter$\texttt{++} \;
    }
  }
  \caption{$solveNQueens(ce, maxIter)$}
\end{algorithm}

\eus Constraint Reasoning Engine or \texttt{CRE} can be used to compute the
non-attack constraint violations, with \texttt{ce.getViolations()} a
call on the \eu API. The \texttt{CRE} can also be used to pick the queen 
that has he most violations in \texttt{getQueenWithMaxViolation()}, and to
look ahead and see what moves are possible for a queen that would remove its
violations in \texttt{getMoves()}.

### 2012-02-24 trunk/results.tex
final updates

A significant driver of the applicability of our work was the
growing need for AUVs mission to be driven by scientific sensor 
data.Thid trend is illustrated by the 
implementation of MBARI's \emph{Gulper} water-samplers \cite{Bird07},
$2$ litre syringe style containers $10$ of which are encased within
the mid-body section of our Dorado vehicle as shown in Fig
\ref{fig:gulper}. However, the original design and intended use of
operation continued to dominated by a naive belief of collecting and
returning water samples as traditional ship-borne methods often
employ, namely using pre-scripted methods, albeit on a highly capable
robot. With the installation of \rx onboard the Dorado with its
capability to synthesize plans and replan on the fly, more imaginative
methods of sample acquisition were targeted.

Our early work focused on observing and sampling within Intermediate
Nephaloid Layers (INLs) \footnote{INLs are fluid sheets of suspended
  particulate matter that originate from the sea floor
  \cite{mcphee-shaw2006}} \cite{ryan10}. The location and scale of
INLs in the water column, which depend upon seasonal and episodic
variability, are not predictable. In turn, they can be identified by
their high turbidity and low chlorophyll fluorescence. Initial
objectives focused on ways to enhance survey results by ensuring as
large a coverage as possible so that INL fields were observed at
tighter spaced transects and when no INL is detected, then transect
spacing in the lawn mower can be incrementally increased. Follow on
work directly targeted sample acquisition within INLs which are often
sparse, patchy in large horizontal scales (in Kms) and small vertical
scales (in meters). Fig. \ref{fig:inl} shows one example from an AUV
transect from 2005. The intent of both observation and sample
acquisition was to interrupt the nominal plan-execution cycle and
inject an action to alter and re-factor the plan dynamically; in one
instance altering the navigation, the other in actuating a sampler.

>>>>

A significant driver of the applicability of our work was the growing
need for AUV missions to be driven by sensor data. This trend is
illustrated by the implementation of MBARI's \emph{Gulper}
water-samplers \cite{Bird07}, $2$ litre syringe style containers $10$
of which are encased within the mid-body section of our Dorado vehicle
as shown in Fig \ref{fig:gulper}. However, the original design and
intended use of operation continued to be dominated by a naive belief
of collecting and returning water samples as traditional ship-borne
methods often employ, namely using pre-scripted methods. With the
installation of \rx onboard the Dorado with its capability sense the
environment and to appropriately synthesize plans and replan on the
fly, more imaginative methods of sample acquisition were targeted.

Our early work focused on observing and sampling within Intermediate
Nephaloid Layers (INLs) \footnote{INLs are fluid sheets of suspended
  particulate matter that originate from the sea floor
  \cite{mcphee-shaw2006}.} \cite{ryan10}. The location and scale of
INLs in the water column, which depend upon seasonal and episodic
variability, are not predictable. In turn, they can be identified by
their high turbidity and low chlorophyll fluorescence. Initial
objectives focused on ways to enhance survey results by ensuring as
large a coverage as possible so that INL fields were observed at
tighter spaced transects and when no INL is detected, then transect
spacing in the lawn mower can be incrementally increased. Follow on
work directly targeted sample acquisition within INLs which are often
sparse, patchy in large horizontal scales (in Kms) and small vertical
scales (in meters). Fig. \ref{fig:inl} shows one example from an AUV
transect from 2005. The intent of both observation and sample
acquisition was to interrupt the nominal plan-execution cycle and
inject an action to alter and re-factor the plan dynamically; in one
instance altering the navigation, the other in actuating a sampler.
